---
title: "clustering_analysis"
format: html
editor: source
---

# Set-up

```{r setup, message = FALSE, warning = FALSE, results = 'hide'}
library(HDAnalyzeR)
library(tidyverse)
library(ggalluvial)
```

# Import data & metadata

```{r, message = FALSE, results = 'hide'}
data_scapis <- import_df("data/processed_data/scapis_joined_preprocessed.rda")
data_igt <- import_df("data/processed_data/igt_joined_preprocessed.rda")
metadata <- import_df("data/processed_data/metadata_preprocessed.rda")
```

# Import LASSO model features

```{r, message = FALSE, results = 'hide'}
lasso_features <- readRDS("results/models_results/lasso_features.RData")
lasso_onlyproteins_features <- readRDS("results/models_results/lasso_onlyproteins_features.RData")
```

# Clustering analysis

## Features vs diseases

### Identify proteomics features

I select features that have more than 50% Importance. In case that the disease has less than 3 features with more than 50% Importance, I select the top 3 features. On the other hand, if a disease has more than 30 features with more than 50% Importance, I select the top 20 features.

```{r}
top_3_protein <- lasso_features |> 
  group_by(Disease) |> 
  top_n(3, wt = Scaled_Importance)
  
proteomics_features_50 <- lasso_features |> 
  filter(Scaled_Importance > 50) |> 
  group_by(Disease) |>
  slice_max(order_by = Scaled_Importance, n = 20) |> 
  ungroup()

proteomics_features_list <- proteomics_features_50 |> 
  bind_rows(top_3_protein) |>
  pull(Variable) |> 
  unique()
```

```{r}
proteomics_dataset <- data_igt |> 
  select(DAid, any_of(proteomics_features_list))
```

### Identify disease features

```{r}
disease_cols <- c("Obesity", "T2D", "IGT", "MS", "MASLD", "Carotid_plaque", "CAC", "Sleep_apnea")
```

```{r}
disease_dataset <- data_igt |> 
  select(DAid, any_of(disease_cols))
```

```{r}
disease_count <- data_igt |> 
  select(DAid, all_of(disease_cols)) |> 
  pivot_longer(cols = -DAid, names_to = "Disease", values_to = "Disease_status") |> 
  filter(Disease_status == 1) |> 
  group_by(DAid) |> 
  summarise(n = n())
```

### Hierarchical clustering

Run clustering based on both predictors and diseases.

They do not cluster nicely no matter the number of clusters (4-12).

> Categorical predictors are not included (Sex, Smoke_status).

```{r}
num_clusters <- 6  # Specify the number of clusters for the disease data

# Disease clustering
disease_dist <- dist(disease_dataset |> select(-DAid), method = "euclidean")
disease_hclust <- hclust(disease_dist, method = "ward.D2")
disease_clusters <- cutree(disease_hclust, k = num_clusters)

disease_cluster_results <- disease_dataset |> 
  select(DAid) |> 
  mutate(disease_cluster = disease_clusters)

# Combined clustering
proteome_dist <- dist(proteomics_dataset |> select(-DAid), method = "euclidean")
proteome_hclust <- hclust(proteome_dist, method = "ward.D2")
proteome_clusters <- cutree(proteome_hclust, k = num_clusters)

proteome_cluster_results <- proteomics_dataset |> 
  select(DAid) |> 
  mutate(proteome_cluster = proteome_clusters)
```

Compare clusters.

```{r}
cluster_results <- disease_cluster_results |> 
  inner_join(proteome_cluster_results, by = "DAid") |> 
  mutate(proteome_cluster = paste0("F", proteome_cluster),
         disease_cluster = paste0("D", disease_cluster),
         DummyCount = 1) |> 
  left_join(disease_count, by = "DAid") |>
  mutate(n = ifelse(is.na(n), 0, n))

cluster_results |> 
  ggplot(aes(y = DummyCount, axis1 = disease_cluster, axis2 = proteome_cluster, color = n)) +
  geom_alluvium(width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Disease", "Features"), expand = c(.05, .05)) +
  labs(y = "") +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```

### KMeans clustering

Run clustering based on both predictors and diseases.

They do not cluster nicely no matter the number of clusters (4-12).

> Categorical predictors are not included (Sex, Smoke_status).

```{r}
num_clusters <- 6  # Specify the number of clusters for the disease data

# Disease clustering
disease_kmeans <- disease_dataset |> 
  select(-DAid) |> 
  kmeans(centers = num_clusters)

disease_cluster_results <- data.frame(DAid = disease_dataset$DAid, 
                                      disease_cluster = disease_kmeans$cluster)

# Combined clustering
proteome_kmeans <- proteomics_dataset |> 
  select(-DAid) |> 
  kmeans(centers = num_clusters)

proteome_cluster_results <- data.frame(DAid = proteomics_dataset$DAid, 
                                       proteome_cluster = proteome_kmeans$cluster)
```

Compare clusters.

```{r}
cluster_results <- disease_cluster_results |> 
  inner_join(proteome_cluster_results, by = "DAid") |> 
  mutate(proteome_cluster = paste0("F", proteome_cluster),
         disease_cluster = paste0("D", disease_cluster),
         DummyCount = 1)

cluster_results |> 
  ggplot(aes(y = DummyCount, axis1 = disease_cluster, axis2 = proteome_cluster)) +
  geom_alluvium(width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Disease", "Features"), expand = c(.05, .05)) +
  labs(y = "") +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```

## Exclude CAC, Carotid Plaque and IGT

### Identify proteomics features

```{r}
top_5_protein <- lasso_features |> 
  filter(Disease != "CAC" & Disease != "Carotid_plaque" & Disease != "IGT" & Disease != "Sleep_apnea") |>
  group_by(Disease) |> 
  top_n(5, wt = Scaled_Importance)
  
proteomics_features_50 <- lasso_features |> 
  filter(Disease != "CAC" & Disease != "Carotid_plaque" & Disease != "IGT" & Disease != "Sleep_apnea") |>
  filter(Scaled_Importance > 50) |> 
  group_by(Disease) |>
  slice_max(order_by = Scaled_Importance, n = 20) |> 
  ungroup()

proteomics_features_list <- proteomics_features_50 |> 
  bind_rows(top_5_protein) |>
  pull(Variable) |> 
  unique()
```

```{r}
proteomics_dataset <- data_igt |> 
  select(DAid, any_of(proteomics_features_list))
```

### Identify disease features

```{r}
disease_cols <- c("Obesity", "T2D", "MS", "MASLD")
```

```{r}
disease_dataset <- data_igt |> 
  select(DAid, any_of(disease_cols))
```

```{r}
disease_count <- data_igt |> 
  select(DAid, all_of(disease_cols)) |> 
  pivot_longer(cols = -DAid, names_to = "Disease", values_to = "Disease_status") |> 
  filter(Disease_status == 1) |> 
  group_by(DAid) |> 
  summarise(n = n())
```

### Hierarchical clustering

Run clustering based on both predictors and diseases.

> Categorical predictors are not included (Sex, Smoke_status).

```{r}
num_clusters <- 6  # Specify the number of clusters for the disease data

# Disease clustering
disease_dist <- dist(disease_dataset |> select(-DAid), method = "euclidean")
disease_hclust <- hclust(disease_dist, method = "ward.D2")
disease_clusters <- cutree(disease_hclust, k = num_clusters)

disease_cluster_results <- disease_dataset |> 
  select(DAid) |> 
  mutate(disease_cluster = disease_clusters)

# Combined clustering
proteome_dist <- dist(proteomics_dataset |> select(-DAid), method = "euclidean")
proteome_hclust <- hclust(proteome_dist, method = "ward.D2")
proteome_clusters <- cutree(proteome_hclust, k = num_clusters)

proteome_cluster_results <- proteomics_dataset |> 
  select(DAid) |> 
  mutate(proteome_cluster = proteome_clusters)
```

Compare clusters.

```{r}
cluster_results <- disease_cluster_results |> 
  inner_join(proteome_cluster_results, by = "DAid") |> 
  mutate(proteome_cluster = paste0("F", proteome_cluster),
         disease_cluster = paste0("D", disease_cluster),
         DummyCount = 1) |> 
  left_join(metadata |> select(DAid, Obesity), by = "DAid")

cluster_results |> 
  ggplot(aes(y = DummyCount, axis1 = disease_cluster, axis2 = proteome_cluster, color = Obesity)) +
  geom_alluvium(width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Disease", "Features"), expand = c(.05, .05)) +
  labs(y = "") +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```

```{r}
do_umap(data_igt |> select(DAid, any_of(proteomics_features_list), -Sex, -Smoke_status),
        metadata |> 
          filter(DAid %in% data_igt$DAid) |> 
          left_join(cluster_results) |> 
          mutate(disease_cluster = factor(disease_cluster)),
        color = "disease_cluster")
```

### KMeans clustering

Run clustering based on both predictors and diseases.

They do not cluster nicely no matter the number of clusters (4-12).

> Categorical predictors are not included (Sex, Smoke_status).

```{r}
num_clusters <- 6  # Specify the number of clusters for the disease data

# Disease clustering
disease_kmeans <- disease_dataset |> 
  select(-DAid) |> 
  kmeans(centers = num_clusters)

disease_cluster_results <- data.frame(DAid = disease_dataset$DAid, 
                                      disease_cluster = disease_kmeans$cluster)

# Combined clustering
proteome_kmeans <- proteomics_dataset |> 
  select(-DAid) |> 
  kmeans(centers = num_clusters)

proteome_cluster_results <- data.frame(DAid = proteomics_dataset$DAid, 
                                       proteome_cluster = proteome_kmeans$cluster)
```

Compare clusters.

```{r}
cluster_results <- disease_cluster_results |> 
  inner_join(proteome_cluster_results, by = "DAid") |> 
  mutate(proteome_cluster = paste0("F", proteome_cluster),
         disease_cluster = paste0("D", disease_cluster),
         DummyCount = 1)

cluster_results |> 
  ggplot(aes(y = DummyCount, axis1 = disease_cluster, axis2 = proteome_cluster)) +
  geom_alluvium(width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Disease", "Features"), expand = c(.05, .05)) +
  labs(y = "") +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```

## Proteins vs clinical metadata

### Identify proteomics features

I select features that have more than 50% Importance. In case that the disease has less than 3 features with more than 50% Importance, I select the top 3 features. On the other hand, if a disease has more than 30 features with more than 50% Importance, I select the top 20 features.

```{r}
top_3_protein <- lasso_onlyproteins_features |> 
  group_by(Disease) |> 
  top_n(3, wt = Scaled_Importance)
  
proteomics_features_50 <- lasso_onlyproteins_features |> 
  filter(Scaled_Importance > 50) |> 
  group_by(Disease) |>
  slice_max(order_by = Scaled_Importance, n = 20) |> 
  ungroup()

proteomics_features_list <- proteomics_features_50 |> 
  bind_rows(top_3_protein) |>
  pull(Variable) |> 
  unique()
```

```{r}
proteomics_dataset <- data_igt |> 
  select(DAid, any_of(proteomics_features_list))
```

### Identify clinical features

```{r}
clinical_cols <- c("Age", "MAP_Mean", "HbA1c", "Chol", "LDL", "HDL", "TG", "ALT", "GGT", "Urate", "Crea", "eGFR", "CRP", "Hb", "WBC", "PLT")
```

```{r}
clinical_dataset <- data_igt |> 
  select(DAid, any_of(clinical_cols))
```

### Hierarchical clustering

Run clustering based on both predictors and diseases.

They do not cluster nicely no matter the number of clusters (4-12).

> Categorical predictors are not included (Sex, Smoke_status).

```{r}
num_clusters <- 6  # Specify the number of clusters for the disease data

# Clinical clustering
clinical_dist <- dist(clinical_dataset |> select(-DAid), method = "euclidean")
clinical_hclust <- hclust(clinical_dist, method = "ward.D2")
clinical_clusters <- cutree(clinical_hclust, k = num_clusters)

clinical_cluster_results <- clinical_dataset |> 
  select(DAid) |> 
  mutate(clinical_cluster = clinical_clusters)

# Proteome clustering
proteome_dist <- dist(proteomics_dataset |> select(-DAid), method = "euclidean")
proteome_hclust <- hclust(proteome_dist, method = "ward.D2")
proteome_clusters <- cutree(proteome_hclust, k = num_clusters)

proteome_cluster_results <- proteomics_dataset |> 
  select(DAid) |> 
  mutate(proteome_cluster = proteome_clusters)
```

Compare clusters.

```{r}
cluster_results <- clinical_cluster_results |> 
  inner_join(proteome_cluster_results, by = "DAid") |> 
  mutate(proteome_cluster = paste0("P", proteome_cluster),
         clinical_cluster = paste0("C", clinical_cluster),
         DummyCount = 1)

cluster_results |> 
  ggplot(aes(y = DummyCount, axis1 = clinical_cluster, axis2 = proteome_cluster)) +
  geom_alluvium(width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Clinical", "Proteome"), expand = c(.05, .05)) +
  labs(y = "") +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```

### KMeans clustering

Run clustering based on both predictors and diseases.

They do not cluster nicely no matter the number of clusters (4-12).

> Categorical predictors are not included (Sex, Smoke_status).

```{r}
num_clusters <- 6  # Specify the number of clusters for the disease data

# Clinical clustering
clinical_kmeans <- clinical_dataset |> 
  select(-DAid) |> 
  kmeans(centers = num_clusters)

clinical_cluster_results <- data.frame(DAid = clinical_dataset$DAid, 
                                       clinical_cluster = clinical_kmeans$cluster)

# Proteome clustering
proteome_kmeans <- proteomics_dataset |> 
  select(-DAid) |> 
  kmeans(centers = num_clusters)

proteome_cluster_results <- data.frame(DAid = proteomics_dataset$DAid, 
                                       proteome_cluster = proteome_kmeans$cluster)
```

Compare clusters.

```{r}
cluster_results <- clinical_cluster_results |> 
  inner_join(proteome_cluster_results, by = "DAid") |> 
  mutate(proteome_cluster = paste0("P", proteome_cluster),
         clinical_cluster = paste0("C", clinical_cluster),
         DummyCount = 1)

cluster_results |> 
  ggplot(aes(y = DummyCount, axis1 = clinical_cluster, axis2 = proteome_cluster)) +
  geom_alluvium(width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Clinical", "Proteome"), expand = c(.05, .05)) +
  labs(y = "") +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```

## Clinical metadata vs diseases

### Identify disease features

```{r}
disease_cols <- c("Obesity", "T2D", "IGT", "MS", "MASLD", "Carotid_plaque", "CAC", "Sleep_apnea")
```

```{r}
disease_dataset <- data_igt |> 
  select(DAid, any_of(disease_cols))
```

### Identify clinical features

```{r}
clinical_cols <- c("Age", "MAP_Mean", "HbA1c", "Chol", "LDL", "HDL", "TG", "ALT", "GGT", "Urate", "Crea", "eGFR", "CRP", "Hb", "WBC", "PLT")
```

```{r}
clinical_dataset <- data_igt |> 
  select(DAid, any_of(clinical_cols))
```

### Hierarchical clustering

Run clustering based on both predictors and diseases.

They do not cluster nicely no matter the number of clusters (4-12).

> Categorical predictors are not included (Sex, Smoke_status).

```{r}
num_clusters <- 6  # Specify the number of clusters for the disease data

# Clinical clustering
clinical_dist <- dist(clinical_dataset |> select(-DAid), method = "euclidean")
clinical_hclust <- hclust(clinical_dist, method = "ward.D2")
clinical_clusters <- cutree(clinical_hclust, k = num_clusters)

clinical_cluster_results <- clinical_dataset |> 
  select(DAid) |> 
  mutate(clinical_cluster = clinical_clusters)

# Disease clustering
disease_dist <- dist(disease_dataset |> select(-DAid), method = "euclidean")
disease_hclust <- hclust(disease_dist, method = "ward.D2")
disease_clusters <- cutree(disease_hclust, k = num_clusters)

disease_cluster_results <- disease_dataset |> 
  select(DAid) |> 
  mutate(disease_cluster = disease_clusters)
```

Compare clusters.

```{r}
cluster_results <- clinical_cluster_results |> 
  inner_join(disease_cluster_results, by = "DAid") |> 
  mutate(disease_cluster = paste0("D", disease_cluster),
         clinical_cluster = paste0("C", clinical_cluster),
         DummyCount = 1)

cluster_results |> 
  ggplot(aes(y = DummyCount, axis1 = clinical_cluster, axis2 = disease_cluster)) +
  geom_alluvium(width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Clinical", "Disease"), expand = c(.05, .05)) +
  labs(y = "") +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```

### KMeans clustering

Run clustering based on both predictors and diseases.

They do not cluster nicely no matter the number of clusters (4-12).

> Categorical predictors are not included (Sex, Smoke_status).

```{r}
num_clusters <- 6  # Specify the number of clusters for the disease data

# Clinical clustering
clinical_kmeans <- clinical_dataset |> 
  select(-DAid) |> 
  kmeans(centers = num_clusters)

clinical_cluster_results <- data.frame(DAid = clinical_dataset$DAid, 
                                       clinical_cluster = clinical_kmeans$cluster)

# Disease clustering
disease_kmeans <- disease_dataset |> 
  select(-DAid) |> 
  kmeans(centers = num_clusters)

disease_cluster_results <- data.frame(DAid = disease_dataset$DAid, 
                                      disease_cluster = disease_kmeans$cluster)
```

Compare clusters.

```{r}
cluster_results <- clinical_cluster_results |> 
  inner_join(disease_cluster_results, by = "DAid") |> 
  mutate(disease_cluster = paste0("D", disease_cluster),
         clinical_cluster = paste0("C", clinical_cluster),
         DummyCount = 1)

cluster_results |> 
  ggplot(aes(y = DummyCount, axis1 = clinical_cluster, axis2 = disease_cluster)) +
  geom_alluvium(width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Clinical", "Disease"), expand = c(.05, .05)) +
  labs(y = "") +
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```